---
title: "HW1"
author: '113078506'
date: "2025-02-19"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document: default
---
## Q1: What is the 5th element in the original list of ages?

```{r}
customers <- read.table(file = "customers.txt", header = TRUE)
customers$age[5]
```

## Q2: What is the fifth lowest age?

```{r}
sorted_ages <- sort(customers$age)
sorted_unique_ages <- unique(sorted_ages)
sorted_unique_ages[5]
```

## Q3: Extract the five lowest ages together

```{r}
sorted_unique_ages[1:5]
```

## Q4: Get the five highest ages by first sorting them in decreasing order first.

```{r}
?sort
desc_ages <- sort(sorted_unique_ages, decreasing = TRUE)
desc_ages[1:5]
```

##Q5: What is the average (mean) age?

```{r}
ages_mean <- mean(sorted_ages)
ages_mean
```

##Q6: What is the standard deviation of ages?

```{r}
sd(sorted_ages)
```

##Q7: Make a new variable called age_diff, with the difference between each age and the mean age

```{r}
age_diff <- sorted_ages - ages_mean
age_diff
```

## Q8: What is the average “difference between each age and the mean age”?

```{r}
mean(abs(age_diff))
```

## Q9: Visualize the raw data as we did in class: (a) histogram, (b) density plot, (c) boxplot+stripchart

```{r}
hist(customers$age)
```

```{r}
plot(density(customers$age))
```

```{r}
boxplot(customers$age, horizontal = TRUE)
stripchart(customers$age, method = "stack", add = TRUE)
```

